Isomorphic Interpreters from Logically Reversible Abstract Machines

### 2012-03-15 iso-int/rc-iso-int.tex
intro

Reversible languages

Main points: 
\begin{itemize}
\item Make the case for reversibility via type isomorphisms
\item Type isomorphisms between finite types are straightforward (but
  tedious)
\item Type isomorphisms between recursive types are quite difficult. This is
  what we focus on.
\end{itemize}
To understand the problem in more detail, consider the problem of converting
a reversible recursive function into a partial isomorphism. In general such a
function takes an input, makes several recursive calls, and produces an
answer. To implement this in a reversible way, the structure of the recursion
and the number of recursive calls need to be exposed. The key trick is to
express tease the function into two components: an outer driver that performs
iteration and an inner state transition function that realizes one
``unfolding'' of the recursion. In program transformation world, this is
called, light-weight fission. Using this idea, one can realize any recursive
function as a partial isomorphism. We show a few examples of the construction
culminating with the most challenging recursive function: a full interpreter.

Previous work on {{langRevT}} \cite{rc2011,infeffects} studied its
properties and various constructions in it. However, for a
computational model, there has been little focus on actually writing
programs directly in {{langRevT}}. This paper aims to remedy that by
presenting a systematic technique of expressing logically reversible
small-step abstract machines in {{langRevT}}. This paper shows that
once we have a logically reversible machine in a notation of our
choice, expressing the machine as an isomorphic interpreter in
{{langRevT}} can be done systematically and does not present any
significant conceptual difficulties.

The constructions in this paper apply only to a class of interpreters
that can be represented by logically reversible small-step operational
semantics. While, in a formal sense, they are related to Abramsky's
``bi-orthogonal automata''\cite{abramsky2005structural}, we will try
to give them a more direct characterization purely in terms of logical
reversibility of the small-step semantics.

>>>>

In recent papers~\cite{rc2011,James:2012:IE:2103656.2103667}, we developed a
pure reversible model of computation that is obtained from the type
isomorphisms and categorical structures that underlie models of linear logic
and quantum computing and that treats information as a linear resource that
can neither be erased nor duplicated. From a programming perspective, our
model gives rise to a pure (with no embedded computational effects such as
assignments) reversible programming language {{langRevT}} based on
\emph{partial isomorphisms}. In more detail, in the recursion-free fragment
of {{langRevT}}, every program witnesses an isomorphism between two finite
types; in the full language with recursion, some of these isomorphisms may be
partial, i.e., may diverge on some inputs.

In this paper, we investigate the practicality of writing recursive programs
in {{langRevT}}. Specifically, we assume that we are giving some reversible
recursive algorithm expressed as a \emph{reversible abstract machine}, and we
show via a number of systematic steps, how to develop a {{langRevT}} program
from that abstract machine. Our choice of starting from reversible abstract
machines is supported by the following observations:
\begin{itemize}
\item it is an interesting enough class of reversible programs: researchers
  have invested the effort in manually designing such machines, e.g., the
  SE(M)CD machine~\cite{Kluge:1999:SEMCD}, and theSECD-H~\cite{lorenz});
\item every reversible interpreter can be realized as such a machine: this
  means that our class of programs includes self-interpreters which are
  arguably a measure of the strength of any reversible
  language~\cite{Yokoyama:2008:PRP,Axelsen:2011:RPC:1987171.1987176,Yokoyama:2007:RPL:1244381.1244404};
\item general recursive programs can be systematically transformed to
  abstract machines: the technique is independent of reversible programs and
  consists of tranforming general recursion to tail recursion and then
  applying \emph{fission} to split the program into a driver and a small-step
  machine~\cite{Danvy:2008:RRN:1813347.1813350,Rendel:2010:ISD:1863523.1863525}.
\end{itemize}

To summarize, we assume we are given some reversible abstract machine and we
show how to derive a {{langRevT}} program that implements the semantics of
the machine. Our derivation is systematic and expressive. In particular,
{{langRevT}} can handle machines with stuck states because it is based on
partial isomorphisms. We illustrate our techniques with simple machines that
do bounded iteration on numbers, tree traversals, and a meta-circular
interpreter for {{langRevT}}.

% To understand the problem in more detail, consider the problem of converting
% a reversible recursive function into a partial isomorphism. In general such a
% function takes an input, makes several recursive calls, and produces an
% answer. To implement this in a reversible way, the structure of the recursion
% and the number of recursive calls need to be exposed. The key trick is to
% express tease the function into two components: an outer driver that performs
% iteration and an inner state transition function that realizes one
% ``unfolding'' of the recursion. In program transformation world, this is
% called, light-weight fission. Using this idea, one can realize any recursive
% function as a partial isomorphism. We show a few examples of the construction
% culminating with the most challenging recursive function: a full interpreter.

% Previous work on {{langRevT}} \cite{rc2011,James:2012:IE:2103656.2103667}
% studied its properties and various constructions in it. However, for a
% computational model, there has been little focus on actually writing programs
% directly in {{langRevT}}. This paper aims to remedy that by presenting a
% systematic technique of expressing logically reversible small-step abstract
% machines in {{langRevT}}. This paper shows that once we have a logically
% reversible machine in a notation of our choice, expressing the machine as an
% isomorphic interpreter in {{langRevT}} can be done systematically and does
% not present any significant conceptual difficulties.

% The constructions in this paper apply only to a class of interpreters
% that can be represented by logically reversible small-step operational
% semantics. While, in a formal sense, they are related to Abramsky's
% ``bi-orthogonal automata''\cite{abramsky2005structural}, we will try
% to give them a more direct characterization purely in terms of logical
% reversibility of the small-step semantics.

%% - write program; derive inverse
%% program~\cite{Gluck:2005,Mu:2002:IFF:648086.757330,conf/aplas/GluckK03,Gluck:2005};
%% we're not interested in this; we want to directly write partial isomorphisms
